%
% chapters/chapter01/introduction.tex
% Chapter 1: Theoretical foundations and domain analysis
%

\chapter{Theoretical foundations and domain analysis}
\label{chap:theory}

\section{The relevance of automatic geotag recognition in textual data}
\label{sec:relevance}

A constant flow of unstructured operational data as a result of the rapid digitalization of urban service delivery processes is being received by public transportation providers (e.g. complaints), emergency response organizations, and municipal agencies daily. These types of textual input contain valuable geographic information about where events occurred or problems were reported (street name, stop on bus/train/rail, intersection) which can be used operationally in real-time with reliable extraction. One of the most important challenges for both natural language processing and geospatial computing today is to identify and structure geographic reference in un-structured, multilingual user generated content.

Smart Cities are a growing global movement. Smart City systems aim to enhance living in cities through a combination of data-based technology and information communications technologies (ICTs) used as part of local government services. Transportation management is one of the most significant areas in which smart city initiatives have potential for large-scale impact. Transit agencies need to operate their fleets reliably; respond quickly to failures in their operations; and adjust the bus routes on an ongoing basis based on changing travel patterns of passengers. Present day transportation management models utilize structured sensor data, or “telemetry,” collected from buses including GPS location tracking of buses; number of passengers counted when entering or exiting buses; and whether buses follow their scheduled run times. While these types of telemetry collect \textit{where} and \textit{when} things happen, it does not explain \textit{why} these events happened. As the result of an event occurring while using public transportation (e.g., missing a stop), citizens report what occurred. For example, when a citizen reports, \foreignlanguage{russian}{«автобус маршрут 48 не остановился на Байтурсынова»} ("route #48 didn't make a stop at Baitursunov"), he/she provides valuable and actionable information about the nature of what occurred: route number; the stop that was skipped; and a geolocated position of the stop that was skipped (and therefore, if linked with city's geographic database system – GIS – could send a dispatch alert to a dispatcher in mere seconds). 

The main reason behind this project was the need to automate the triaging process by using City Transportation Systems (CTS), which provides the public transportation service in Astana, Kazakhstan. CTS receives around 3\,000 monthly complaints about the city’s public transportation via three channels; a mobile app, a website portal and a call centre. All these complaints contain an open-ended description field (description) where customers report transportation problems with the city's public transportation system, including but not limited to missing bus stops, irregular routes, bad driver behavior and/or failed payments. At present, all complaints are processed by human operators who have to manually identify the relevant vehicle, stop and road associated with the complaint after reading it and then route the complaint to the applicable dispatcher. The manual processing of complaints is labor intensive; one single operator cannot process more than 60--80 complaints per hour, so when there is a large volume of complaints it may take up to 2-4 hours for them to reach the correct authority.

Automating the process of extracting the geotag from each complaint eliminates this "bottleneck". The triage/routing step is completely automatic if an NLP model is reliable to extract the Vehicle Identifier, Stop Name & Street Reference from each complaint. In addition, there are other substantial value-added analytical uses for the extracted geotags:

\begin{itemize}
  \item \textbf{Identifying hotspots.} By analyzing the geographical grouping of complaints you can easily see which stops/intersections have higher than normal incident rates, thus allowing you to proactively maintain these areas as well as reschedule routes accordingly.
  \item \textbf{Analyzing trends at the route level.} By aggregating route numbers on a day-by-day basis you will be able to detect patterns of reliability degradation prior to when they become evident in your overall performance statistics.
  \item \textbf{Balancing dispatcher workloads.} By associating complaints with specific geographic zones (stops) you can better distribute the load among operators.
  \item \textbf{Overlaying GIS and visualizing data.} Once geotags have been successfully resolved against your City's stop/street database, you can create real-time situational awareness displays used by transportation management to help them visualize and understand current events.
  \item \textbf{Evaluating policy.} The before/after comparisons of the geotagged data associated with complaints can measure the effectiveness of the changes made to routes, investments in infrastructure or training programs for drivers
\end{itemize}

The nature of the CTS Complaint Corpus is difficult to handle due to its combination of different features that make it particularly challenging to be handled by conventional NLP pipelines. Firstly, these complaint texts contain code-switching - the switching from one language into another within or between sentences. Secondly, the informal tone used in these complaint texts contains informal language elements such as abbreviations (e.g., «м/а» = "route / bus"), spelling mistakes, and colloquialisms. In addition, there are large differences between the six geographic entity types, as well as their linguistic composition, occurrence rate and extraction difficulties; thus forming a heterogenous NER-task where no single modeling approach can be optimal for all entity-types at once. Finally, the data is very imbalanced on the classification level. The most frequent type (\texttt{STOP\_NAME}) appears much more often (more than 10 times) than the least frequent type (\texttt{STREET\_INTERSECTION}). Therefore a maximal likelihood-training will have many problems when trying to train the minority-class


The natural language processing (nlp) task most relevant to solving this problem is named Entity Recognition (NER). NER frames geotag extraction as a supervised sequence labeling problem in which each token within the input is assigned a semantic label that represents its class of entities and position within an entity span. The accuracy, robustness and computational efficiency of the chosen architecture for NER will directly determine whether or not it will be feasible to create an automated pipeline to recognize geotags for the CTS. This thesis develops and compares four fundamentally different architectures — fine-tuned multilingual transformer encoders; a Russian-specific BERT model; a multilingual embedding-based model; and large language models deployed via few-shot prompting — to identify which strategy or combination of strategies best meets the requirements of the operational needs of the CTS geotag extraction task.

The current research is based on an earlier study ~\cite{ussenbayeva2026heterogeneous}, which revealed the diverse characteristics of NER within transport domain, and demonstrated how aggregated evaluation metrics may hide some differences per category. In addition to this prior work, the present thesis also develops a wide-ranging comparison of various models and offers concrete guidelines for the selection and deployment of NER models in the same type of environments as those mentioned above (low resource bilingual).



\section{General characteristics of geographic named entities and toponyms in transport-domain texts}
\label{sec:geo_entities}

Geographic named entities (GNE's) that occur naturally in languages have many levels of abstraction: 
from large geographic units (countries, states/provinces, cities), 
to smaller ones (neighborhoods/districts, zip codes/postal zones), 
and then down to small, operational items (street names, building numbers, transit stop locations). 
In complaints about transportation issues, the GNE's most important for the operations are also very granular and specific to the domain; and they differ greatly from what is typically referred to as "spatial reference" in traditional NER benchmark models like CoNLL-2003, which groups all spatial references under one class called \texttt{LOC}.


Geotagged Complaints for Public Transport Operations Have Three Distinct Operational Uses 
\textit{ Complaint geotags are identification tags } (e.g. route number, vehicle identifier, license plate) that associate a complaint with an identifiable piece of equipment used in the field, which is stored in the fleet's management database. \textit{ Complaint geotags are also spatial tags } (e.g. name of stop at which incident occurred; name of road on which incident occurred), which allow a complaint to be associated with a particular location, such as one contained within the City's Geographic Information System (GIS) layers. Finally, \textit{composite tags} (i.e., a combination of two or more spatial references combined into a single structured expression, e.g. "Intersection of Main St & Elm") provide even greater precision than spatial tags when referencing a particular location. As each type of geotag has its own unique structure, annotation characteristics and difficulties in extracting data


Due to corpus analysis and the domains' necessities; Six different types of entities were identified as necessary for the CTS geotag extraction task:

\begin{description}[style=multiline,leftmargin=4.5cm,labelwidth=4.5cm]
  \item[\texttt{ROUTE\_NUM}]           Public transit route number. Examples: \foreignlanguage{russian}{«Маршрут 47»}, \foreignlanguage{russian}{«автобус №~64»}, \foreignlanguage{russian}{«48 бағыт»}. These identifiers typically appear as short numeric spans, optionally preceded by a route keyword.
  \item[\texttt{BUS\_ID}]              Vehicle identifier or fleet (garage) number assigned to a specific bus. Examples: \foreignlanguage{russian}{«гаражный номер 3056»}, \foreignlanguage{russian}{«борт 10815»}. Unlike route numbers, bus IDs identify the physical vehicle rather than the service.
  \item[\texttt{PLATE\_NUM}]           Vehicle licence plate number. Examples: \foreignlanguage{russian}{«КЛС 012»}, \foreignlanguage{russian}{«164 BW 01»}. Plates may combine Cyrillic and Latin characters in various regional format conventions.
  \item[\texttt{STOP\_NAME}]           Named transit or bus stop. Examples: \foreignlanguage{russian}{«аялдама Қобыланды»}, \foreignlanguage{russian}{«остановка Сарыарка»}, \foreignlanguage{russian}{«ТРЦ Керуен»}. Stop names are the most frequent entity type and appear in both Kazakh and Russian forms.
  \item[\texttt{STREET\_NAME}]         Named street, avenue, boulevard, or road. Examples: \foreignlanguage{russian}{«проспект Тауелсиздик»}, \foreignlanguage{russian}{«улица Абая»}, \foreignlanguage{russian}{«Кабанбай батыра даңғылы»}. Street names are morphologically inflected in context.
  \item[\texttt{STREET\_INTERSECTION}] Geographic intersection of two or more streets, expressed as a structured multi-word span. Examples: \foreignlanguage{russian}{«угол Абая и Сейфуллина»}, \foreignlanguage{russian}{«пересечение Тауелсиздик и Республики»}. This is the rarest and structurally most complex entity type.
\end{description}

Table~\ref{tab:entity_typology} summarises the structural characteristics of the six entity categories, including their linguistic nature, typical span length in tokens, and the principal sources of extraction difficulty.

\begin{table}[H]
\centering
\caption{Structural characteristics of entity categories in the CTS corpus (adapted from~\cite{ussenbayeva2026heterogeneous}).}
\label{tab:entity_typology}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{llp{2.2cm}p{3.8cm}}
\toprule
\textbf{Entity Type} & \textbf{Nature} & \textbf{Typical Span (tokens)} & \textbf{Principal Extraction Difficulty} \\
\midrule
\texttt{ROUTE\_NUM}           & Numeric / identifier  & 1--4   & Boundary ambiguity with time codes and QR identifiers \\
\texttt{BUS\_ID}              & Numeric / identifier  & 1--3   & Semantic confusion with \texttt{PLATE\_NUM} \\
\texttt{PLATE\_NUM}           & Alphanumeric          & 1--3   & Mixed Latin-Cyrillic format variation \\
\texttt{STOP\_NAME}           & Nominal / toponym     & 1--8   & Abbreviation, bilingual lexical variation \\
\texttt{STREET\_NAME}         & Morphological         & 2--6   & Case inflection, code-switching boundary \\
\texttt{STREET\_INTERSECTION} & Composite / structured & 5--12 & Internal compositional structure; rare training examples \\
\bottomrule
\end{tabular}
\end{table}

\texttt{ROUTE\_NUM} and \texttt{STOP\_NAME} appear to be the most frequent entity types as they can be found on about 72\,\% and 68\,\% of complaints, respectively. The rarest type is Street_intersection which appears in less than 3\,\% of cases. Structurally, \texttt{STREET\_INTERSECTION} is the most complicated as it has a span of 5--12 tokens including conjunction words (\foreignlanguage{russian}{\textit{угол}}, \foreignlanguage{russian}{\textit{пересечение}}, \foreignlanguage{russian}{\textit{қиылысы}}) with both distinct street name forms with grammatical case marker, and optionally contains additional information such as direction or time.

Transportation domain toponyms have many differences to general language toponyms which make it difficult to do the automatic toponym extraction. 
\textit{Context dependence:} The same word e.g. “Republic” can be used as a reference point (e.g. bus stop name, street name, organization) based on contextual references. 
\textit{Abbreviation:} Stop names and street types are often shortened when written in complaint texts (e.g. \foreignlanguage{russian}{«ул.»} for \foreignlanguage{russian}{«улица»}, \foreignlanguage{russian}{«пр.»} for \foreignlanguage{russian}{«проспект»}, \foreignlanguage{russian}{«м/а»} for \foreignlanguage{russian}{«маршрут/автобус»}). 
\textit{ Orthographical Variants:} Both Russian and Kazakh languages use different forms of writing the same stop name; e.g. \foreignlanguage{russian}{«Байтурсынова»} (Russian) and \foreignlanguage{russian}{«Байтурсынов даңғылы»} (Kazakh) within the same document collection and thus require a model to generalize across those variations. 
\textit{Identifier Overlap:} BUS\_ID and PLATE\_NUM both look like the same type of token, i.e. small number of characters alphanumeric strings but have limited contextual support, resulting in high overlap rates in models that primarily depend on surface form.


Token distributions are very unbalanced. Analysis of dataset shows that the \texttt{O} (Outside) class comprises about 85\,\% of the total tokens, while entity tokens represent fewer than 14\,\%. Between these two entity token types, \texttt{STOP\_NAME} and \texttt{ROUTE\_NUM} are present in around 66\,\% of all entity instances while \texttt{STREET\_INTERSECTION}has fewer than one percent. As such, strong class weighting methods in training objectives, data augmentation methods through sampling techniques, and macro-average evaluation metrics which will assign equal weight to each entity category regardless of its occurrence rate are required. 



\section{Modern approaches to automatic named entity recognition}
\label{sec:modern_approaches}

Named entity recognition (NER) has been defined by three main paradigms in the last thirty years, with each paradigm offering different balances between coverage of language, the requirement for labelled training data, processing costs, and NER performance under low resource or code switching conditions.

\subsection{Rule-based and statistical methods}
\label{sec:rulebased}

The first NER systems were built solely with hand-crafted rule sets. A dictionary of recognized named entities (gazetteer) along with a collection of regular expressions to match standardized writing formats for named entities (such as capitalization, punctuation and number sequences), along with manual coding of contextual knowledge based on these structures formed the base for systems such as ANNIE and LingPipe. While regex based methods can be very accurate for structurally defined identifiers that have been standardized in some way (e.g., license plates; route numbers; postal codes), the method has low recall rates when it does not include all possible variants within its definition.

\textbf{Hidden Markov Models (HMM)}  extended Rule-Based Systems with a Probabilistic Sequence Model. In an HMM based NER System, the observed Output is the Token Sequence and the Hidden States are the BIO Labels. The models jointly maximize:
\begin{equation}
  P(\mathbf{y}, \mathbf{x}) = \prod_{i=1}^{n} P(y_i \mid y_{i-1}) \cdot P(x_i \mid y_i)
\end{equation}
Where $P(y_i \mid y_{i-1})$ represents Transition Probability from one label to another Label, and $P(x_i \mid y_i)$ is the Emission Probability for Token $x_i$ as it relates to label $y_i$. These optimal Label Sequences can be Decoded by using the Viterbi Algorithm. An HMM does model First Order Dependencies in Labels but will not allow Long Range Contextualization nor arbitrary Feature Functions.


\textbf{Conditional Random Fields (CRF)~\cite{lafferty2001crf}} provided an alternative to HMMs as they modeled the \textit{conditional} probability $P(\mathbf{y} \mid \mathbf{x})$, in place of modeling the joint. Linear chain CRF are defined as follows:
\begin{equation}
  P(\mathbf{y} \mid \mathbf{x}) = \frac{1}{Z(\mathbf{x})} \exp\!\left(\sum_{i=1}^{n} \sum_{k} \lambda_k f_k(y_{i-1}, y_i, \mathbf{x}, i)\right)
\end{equation}
where $f_k$ denote feature functions, $\lambda_k$ represent learned weights, and $Z(\mathbf{x})$ denotes the normalization factor.
Feature functions in linear chain CRFs may be used to define arbitrary features from input sequences (e.g., prefix/suffix patterns, capitalization, membership in dictionaries). Richly engineered CRFs were able to hold top NER performance for almost a decade.


\textbf{BiLSTM-CRF} represents the connection between traditional Statistical Methods and New Neural Models. Bidirectional LSTMs take advantage of both left-to-right and right-to-left contextual information when encoding each token as follows:

\begin{equation}
  \overrightarrow{\mathbf{h}}_i = \text{LSTM}_\text{fwd}(e_i, \overrightarrow{\mathbf{h}}_{i-1}), \quad
  \overleftarrow{\mathbf{h}}_i = \text{LSTM}_\text{bwd}(e_i, \overleftarrow{\mathbf{h}}_{i+1})
\end{equation}
\begin{equation}
  \mathbf{h}_i = [\overrightarrow{\mathbf{h}}_i; \overleftarrow{\mathbf{h}}_i]
\end{equation}
Both hidden states are concatenated into one single 
hidden state $\mathbf{h}_i$. This hidden state is fed into a linear layer that produces the label logit scores. The logit scores are then decoded collectively over all tokens by the CRF layer. The CRF ensures globally consistent sequences of labels (for example, an "I-" tag will always follow a "B-" tag of the same part-of-speech category).
Prior to being surpassed by Transformer models, BiLSTM-CRF was able to achieve the highest score in the named-entity recognition task on CoNLL-2003; it is still used today as a strong competitor to many other models, especially those that have limited training data.

Rule-based and statistical models lack generalizability to morphologically complex languages or low resource languages due to their dependency on manually engineered features.


\subsection{Transformer-based encoders (BERT, RoBERTa, XLM-R)}
\label{sec:transformers}

The transformer architecture~\cite{vaswani2017attention} included the self-attention model, which computes context representation by considering the entire input at once through simultaneous position-wise attention. Therefore for an arbitrary input:
\begin{equation}
  \text{Attention}(Q, K, V) = \text{softmax}\!\left(\frac{QK^\top}{\sqrt{d_k}}\right) V
  \label{eq:attention}
\end{equation}
there are Query Matrix $Q$, Key Matrix $K$ and Value Matrix $V$. We also know that the Key Dimension $d_k$. Each Head ($H$) in multi head attention is applied separately and then the outputs are concatenated together as follows: 
\begin{equation}
  \text{MultiHead}(Q, K, V) = [h_1; h_2; \ldots; h_H] W^O, \quad h_i = \text{Attention}(QW_i^Q, KW_i^K, VW_i^V)
\end{equation}
The method allows a single token to pay attention to every other token within the sequence regardless of position; it eliminates the vanishing gradient problem that had limited how long recurrent networks could process sequences.

\textbf{BERT~\cite{devlin2019bert}}, was based upon the concept of the "pre-training" followed by "fine-tuning" paradigm used for NLP. In its "pre-training", there are two unsupervised training objectives applied to large amounts of unlabelled data. The first is Masked Language Modeling (MLM), where 15 \,\% of all input tokens are randomly masked; this causes the model to be trained on predicting the actual token from the context.


\begin{equation}
  \mathcal{L}_\text{MLM} = -\sum_{i \in \mathcal{M}} \log P(x_i \mid \mathbf{x}_{\setminus \mathcal{M}})
\end{equation}
The model is trained using a variant of the Masked Language Modelling objective. Where $\mathcal{M}$ is the set of masked positions. 
Next Sentence Prediction (NSP) objectives trains the model to predict whether two sentences (sentence segments) can occur consecutively in the original document. 
After pre-training, a small task specific-head (a single linear layer for token classification) is added to the top of the model and then the entire model is fine-tuned on labeled data from the task. 
Bert-base achieves an f-score of 91.3 on CoNLL-2003 and significantly exceeds baseline performance based on a bi-lstm crf. 


\textbf{RoBERTa} improved over Bert through removal of the next sentence prediction objectives (this was later found to be detrimental to future models), increased training batch size (from 256 to 8000), extended the amount of training time (bert used 16gb of text while robertabase used 160gb of text) and a dynamic masking method that produced new mask patterns each epoch. 
This provided better quality and less variability in representation; thus, robertalarge achieved an f-score of .93 on CoNLL-2003 ~\cite{keraghel2024survey}.


\textbf{XLM-RoBERTa (XLM-R)~\cite{conneau2020xlmr}}, expanded roberata's pre-train process into 100 languages, utilizing 2.5~tb of common crawl text after filtering. 
The shared subword vocabulary of 250\,000, as defined by Sentencepiece, allowed xlm-r to leverage cross lingual knowledge transfer; thus similar semantic phrases across different languages clustered in their respective embedding spaces. Xlm-r base uses 12 layers of transformers, 768 hidden dimensions, 12 attention heads and approximately 125m parameters. 
Its ability to support both Kazakh and Russian in a singular vocabulary, along with its other properties make it highly suitable for the code switching nature of the cts dataset.
A formal evaluation of xlm-r relative to m-bert has shown that xlm-r performs significantly better than m-bert at NER on all 16 test languages, and also performs relatively well on morphologically complex Slavic and Turkic languages.~\cite{conneau2020xlmr}.


The XLM-R encoder for the NER task is enhanced using an additional token classification head:
The classification head consists of a single linear transformation $W \in \mathbb{R}^{768 \times K}$ maps each contextualized word representation $\mathbf{h}_i \in \mathbb{R}^{768}$ into a vector of class logits. Then, after applying a softmax activation function, we obtain the probability distribution over all possible labels:
\begin{equation}
  p(y_i = c \mid \mathbf{x}) = \frac{\exp(\mathbf{h}_i^\top \mathbf{w}_c)}{\sum_{j=1}^{K} \exp(\mathbf{h}_i^\top \mathbf{w}_j)}
  \label{eq:softmax_ch1}
\end{equation}
Here $K = 13$ corresponds to the total amount of different labels in our dataset (i.e., 6 entity types $\times$ 2 BIO prefixes + $O$).

\subsection{Embedding-based pipelines (E5, multilingual models)}
\label{sec:embeddings}

A parallel body of work generates dense text embeddings optimized to model semantic similarity and support information retrieval tasks via either an instruction-tuned or contrastive learning paradigm as opposed to a masked language modeling paradigm. For example, these models generate representations where semantically similar texts are similarly situated within the representation space regardless of their superficial characteristics. 

\textbf{Multilingual-E5 (mE5)~\cite{wang2023e5}} represents a set of models based upon the XLM-RoBERTa architecture. mE5 was trained on approximately one billion multilingual text pair examples utilizing a multi-stage process that first utilized weak supervision from noisy parallel data followed by fine tuning against high quality supervising data derived from MS MARCO, Natural Questions and TriviaQA. The primary objective of the training function is to provide a contrastive loss that will pull together the representation(s) of each pair of semantically equivalent text while pushing apart the representation(s) of non-equivalent pairs:
\begin{equation}
  \mathcal{L}_\text{contrastive} = -\log \frac{\exp(\text{sim}(\mathbf{q}, \mathbf{d}^+) / \tau)}{\exp(\text{sim}(\mathbf{q}, \mathbf{d}^+) / \tau) + \sum_{j} \exp(\text{sim}(\mathbf{q}, \mathbf{d}^-_j) / \tau)}
\end{equation}
where $\mathbf{q}$ is a query embedding, $\mathbf{d}^+$ is a positive document, $\mathbf{d}^-_j$ are in-batch negatives, and $\tau$ is a temperature parameter.

While originally intended for use at the sentence level, mE5's contextual token representations may be used at the token level using a token based projection instead of pooling tokens together by their sentences. For the purposes of this thesis the \texttt{multilingual-e5-base} (278M parameters) was chosen because it struck a balance between high-quality vector representations and reasonably good computational performance.

Petasis and Karkaletsis ~\cite{petasis2024greek} show that E5-Large Vector Representations combined with an ensemble of traditional machine learning classifiers (XGBoost, k-Nearest Neighbor, Logistic Regression), through Soft Voting, achieves Accuracy of .92 and F1 of .91 on a low resource Greek Named Entity Recognition Task, substantially better than several state-of-the-art Transformer Fine-Tuning Baselines. These results indicate that strong pre-trained vector representations can be sufficient to overcome limitations associated with small amounts of labeled training data, which will likely be true for many of the rare named entities found in the CTS Corpus.

\subsection{Large language models for few-shot NER}
\label{sec:llm_ner}

Recent years have seen the rise of instruction-tuned large language models (LLMs)—models with hundreds of billions of parameters that were trained on vast multi-task data sources and enable task adaptation through natural language prompts instead of gradient-based weight updates. The few-shot paradigm ~\cite{yang2023survey} provides the model with a small number of task-specific examples in the prompt and asks it to generalize to new inputs without any weight updates.

\textbf{In-context learning for NER}frames the extraction task as a conditional text generation problem: given the system prompt, few-shot examples, and the input complaint text, the model generates a JSON object mapping entity types to extracted spans. This approach does not require labelled training data beyond the couple of shots used as examples, which makes it very attractive for entity type with very limited training instances.

\textbf{Qwen2.5~\cite{yang2024qwen}}, is an open-wight llm series (0.5b – 72b parameter count), developed by Alibaba. The 7B-parameter variant supports 29 + languages including Russian and Kazakh, and showcase strong structured output generation using instruction tuning on supervised fine-tuning (SFT) and Reinforcement Learning from human feedback (rlhf) datasets. Instruction tuning is crucial for the JSON structure of the NER output: a model must not only identify entity spans but also reliably format the output as valid JSON with correct field names.

\textbf{ Performance comparison of different LLMs in terms of NER performance.} Keraghel et al.~\cite{keraghel2024survey} report that GPT-4 in zero-shot mode has a value of F1=0.83 on conll-2003. This value reduces to within 10 percentage points of finetuned RoBERTa-large (F1=0.93). Llama3.1 and mistral have lower values of performance (F1\,$\approx$\,0.60--0.70) on this benchmark, especially on entities with limited lexical cues. four LLMs are evaluated under identical few-shot conditions in this thesis to distill the effect of capacity and multilingual pretraining coverage.

\textbf{The design of prompts} for the NER task using an LLM will influence how well it can perform. The entity types included in the prompt define what entities the system should look for in addition to defining extraction guidelines for those entities as natural language. Examples provided to the system were chosen so that all entity types would be represented and specifically, so that the least frequent classes (\texttt{STREET\_INTERSECTION}, \texttt{STREET\_NAME}) would be represented because it is likely these types would not have been encountered by the model during pre-training enough times to have developed knowledge about them.

\section{Using machine learning and natural language processing for multilingual text analysis}
\label{sec:multilingual}

Multimodal NLP seeks to develop Language Technology for multiple languages simultaneously; especially for those who do not have a large amount of data available in standard benchmark datasets. The Kazakh-Russia bilingual environment poses its own set of unique challenges for developing the system architecture.

\textbf{Code-switching} occurs when there is intra-sentential mixing of two or more languages per single utterance. Approximately 67 \,\%  of complaint records within the CTS dataset include at least one sentence-internal language shift. An example is as follows: \foreignlanguage{russian}{«м/а 48, уақыты 09:30, остановка Тауелсиздик бағытында жоқ»} (route 48, time 09:30, no stop in the direction of Tauelsizidik), where transit terminology in Russian is combined with temporal and directional phrases in Kazakh. Because code-switching violates the monolingual assumptions inherent to single language training corpora, it can result in out-of-vocabulary errors or diminished representations for tokens in the less represented language.

\textbf{Kazakh morphological complexity.} Kazakh is an agglutinative language in which grammatical information (such as case, number, possession, tense) is expressed using a string of affixes attached to the root of the word. One noun can exhibit up to nine different surface forms depending upon the grammatical case being used. Therefore, if a geotag is required to recognize a "stop" named \foreignlanguage{russian}{«Байтурсынов»}, then it could occur in any of the following forms: \foreignlanguage{russian}{«Байтурсыновта»} (locative: at Baitursunov); \foreignlanguage{russian}{«Байтурсыновқа»} (dative: toward Baitursunov); or \foreignlanguage{russian}{«Байтурсыновтан»} (ablative: from Baitursunov). Although sub-word tokenization (using Sentence Piece BPE) does help with representing inflectionally complex words as composed of root and suffix pieces, the BIO tags still need to be assigned based upon whole word units to ensure annotation consistency.

\textbf{Russian morphological complexity.} As a fusional language, Russian has six cases and two genders; gender affects how adjectives agree with nouns. Streets in Russia are named using varying cases based on the syntactical position they occupy: \foreignlanguage{russian}{«проспект Абая»} (nominative) versus \foreignlanguage{russian}{«на проспекте Абая»} (prepositional) versus \foreignlanguage{russian}{«по проспекту Абая»}. Regardless of whether there is a case-inflectioned preposition or not, each entity includes all of the full canonical parts of the street name.

\textbf{Low-resource asymmetry.} Kazakh is low resource relative to Russian within most multilingual pre-training corpora. For example, while Russian is among the top ten most represented languages in CommonCrawl, the representation of Kazakh is much smaller. The result of this difference is that multilingual models will likely have less capacity to represent morphological features specific to Kazakh resulting in suboptimal subword representations when representing Kazakh tokens that co-exist with other languages within an input.

\textbf{Cross-lingual transfer learning} learning refers to how a multilingual model uses information it has learned about one language in order to help with the improvement of its performance when using another language. A key part of this is that there is an implicit alignment across languages within the shared subword vocabulary of XLMRoBERTA between the representations of semantically-related token types. The representation of \foreignlanguage{russian}{«маршрут»} (Russian: route) will be similar to \foreignlanguage{russian}{«бағыт»} (Kazakh: route) even though they have no common surface form. This cross-language alignment is also important for CTS since the same entity type could be expressed in either language

Iqbal et al.~\cite{iqbal2024multilingual} demonstrated that multilingual BERT (mBERT) achieved a macro F1 score of 0.98 on an augmented multilingual NER dataset and exceeded the monolingual RoBERTa (F1\,=\,0.884). Iqbal et al. attribute mBERT's superiority over monolingual RoBERTa to cross-lingual transfer rather than solely due to increased model capacity, thereby confirming that multilingual pre-training provides benefit even for individual-language NER tasks. Kassab et al. ~\cite{kassab2025russian} found that DeepPavlov RuBERT still had the greatest accuracy as a single Russian-language NER model and had a 28\,\% increase in precision over commercial competitors due to its pre-training specifically on large amounts of Russian web content. However, RuBERT is inherently incapable of being used as a model for processing data containing Russian and Kazakh unless it receives substantial additional training.


Tao et al.~\cite{tao2022socialNER} show that the use of BERT + BiLSTM + CRF and NLP specific pre-processing achieve an F1 score of 0.93 on a Chinese language dataset, with geographical annotations from social media.
They also demonstrated that using contextual embeddings together with preprocessing for informal registers can significantly increase the accuracy of predictions based on noisy, user-generated content. The findings by Tao et al. are directly applicable to the CTS corpus because it has several similarities to social media text (informal register, abbreviations, inconsistent punctuation, etc.) as well as embedded entity references within unstructured prose.


Geoparsing --- combining geographic entities into their respective latitude/longitude coordinates through both geographic entity recognition (GER) and coordinate resolution—is an extension of named-entity recognition (NER). Hu et al. ~\cite{hu2023geoparsing}, assessed 27 geoparsers from 26 publicly available data sets. The authors reported that BERT-type models achieved the best F1 scores (of all the evaluated models) for the task of identifying geographic names in social media texts (i.e., TopoBERT F1\,=\,0.854), while the performance of traditional geoparsers (e.g., Stanford NER) fell off significantly when tested against very noisy inputs. Moncla et al. ~\cite{moncla2022toponyms}, developed methods using CamemBERT and XLM-RoBERTa to improve upon prior work by resolving ambiguities associated with topographic terms; these improvements were demonstrated on multiple French geographic data collections. The authors further reported that the use of neural classifiers on top of the embedding space provided by transformer architectures could resolve spatial ambiguity issues where other systems failed due to reliance on rules. 

Sun et al.~\cite{sun2023chinesectre}, illustrated that incremental training of BERT (on a domain-specific corpus) to support a Chinese social media language environment, when followed by a combination of conditional random field (CRF) and error correction processing (ChineseCTRE), produced a significant increase in the recall of geographic NER (4-5 percent greater than baseline BERT), and therefore indicated that domain-adaptive pre-training of large-scale pre-trained models can be an effective method for improving the performance of minority classes within specific domains.

\section{Comparison of NER architectures for low-resource and code-switched (Russian--Kazakh) settings}
\label{sec:arch_comparison}

Table~\ref{tab:litreview} summarises the 21 key related works surveyed for this thesis, spanning rule-based, statistical, encoder, embedding, and LLM-based NER approaches. The works are presented in the order introduced in the thesis presentation.

{\scriptsize
\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{1.2}
\begin{longtable}{
|>{\centering\arraybackslash}p{0.38cm}
|>{\raggedright\arraybackslash}p{2.6cm}
|>{\raggedright\arraybackslash}p{3.9cm}
|>{\raggedright\arraybackslash}p{3.0cm}
|>{\raggedright\arraybackslash}p{4.5cm}|}
\caption{Summary of key related works on NER and geotag extraction.}
\label{tab:litreview}\\
\hline
\textbf{No.} & \textbf{Author(s) \& Year} & \textbf{Model(s) Used} & \textbf{Key Metric} & \textbf{Key Finding} \\
\hline
\endfirsthead
\multicolumn{5}{l}{\footnotesize\textit{Table~\ref{tab:litreview} continued from previous page}}\\
\hline
\textbf{No.} & \textbf{Author(s) \& Year} & \textbf{Model(s) Used} & \textbf{Key Metric} & \textbf{Key Finding} \\
\hline
\endhead
\hline
\multicolumn{5}{r}{\footnotesize\textit{Continued on next page}}\\
\endfoot
\hline
\endlastfoot
1 & Yang et al.\ (2023)~\cite{yang2023survey} & BERT, GPT-2/3/4, T5, LLaMA, BLOOM & Survey (no dataset) & Covers evolution from BERT to ChatGPT across pretraining objectives and NLP tasks \\
\hline
2 & Keraghel et al.\ (2024)~\cite{keraghel2024survey} & BiLSTM-CRF, BERT, RoBERTa, GPT-3/4, LLaMA & RoBERTa F1\,=\,0.93; GPT-4 zero-shot\,=\,0.83 & Fine-tuned encoders dominate supervised NER; LLMs excel in few-shot settings \\
\hline
3 & Hu et al.\ (2023)~\cite{hu2023geoparsing} & 27 tools incl.\ TopoBERT, spaCy, Stanford NER & TopoBERT F1\,=\,0.854 & BERT-based models best on social media; classical tools degrade sharply on noisy text \\
\hline
4 & Iqbal et al.\ (2024)~\cite{iqbal2024multilingual} & mBERT, RoBERTa-small/large, BERT-base/large & mBERT macro F1\,=\,0.98 & Multilingual BERT outperforms monolingual RoBERTa on augmented multilingual NER \\
\hline
5 & Kassab et al.\ (2025)~\cite{kassab2025russian} & DeepPavlov RuBERT, Natasha, Stanza, spaCy & +28\% precision over single RuBERT & Semi-automated multi-model voting reduces manual annotation effort drastically \\
\hline
6 & Petasis \& Karkaletsis (2024)~\cite{petasis2024greek} & E5-large + XGBoost, KNN, LogReg (soft voting) & Accuracy\,=\,0.92; F1\,=\,0.91 & E5 embeddings + ML ensemble beats fine-tuned transformers on low-resource NER \\
\hline
7 & Wang et al.\ (2023)~\cite{wang2023e5} & mE5-small/base/large/large-instruct & BEIR score\,=\,64.4 (large-instruct) & Two-stage contrastive training yields strong multilingual retrieval embeddings \\
\hline
8 & Yang et al.\ (2024)~\cite{yang2024qwen} & Qwen2.5 (0.5B--72B, Base, Instruct, MoE) & MMLU, HumanEval benchmarks & Instruction tuning improves structured JSON output; open-weight models cost-effective \\
\hline
9 & Mai et al.\ (2024)~\cite{mai2024geollm} & Mistral, Baichuan2, Llama~2, Falcon + geo-KB & Improved geoparsing precision/fairness & LLM fused with geographic knowledge bases improves geoparsing for underrepresented regions \\
\hline
10 & Tarannum et al.\ (2026)~\cite{tarannum2026address} & mBERT, RoBERTa-small, BERT-large/base & mBERT macro F1\,=\,0.98 & Multilingual BERT outperforms monolingual RoBERTa on heterogeneous address data \\
\hline
11 & Wei et al.\ (2024)~\cite{wei2024albert} & ALBERT-Conv1D-BiLSTM-CRF & P\,=\,0.981; R\,=\,0.967; F1\,=\,0.974 & Fusing global (ALBERT) and local (Conv1D) features improves historical toponym NER \\
\hline
12 & Moncla et al.\ (2022)~\cite{moncla2022toponyms} & CamemBERT, mBERT, XLM-RoBERTa + classifier & CamemBERT F1\,$\approx$\,0.92 & Neural classifiers on contextual transformer embeddings resolve toponym ambiguity \\
\hline
13 & Tao et al.\ (2022)~\cite{tao2022socialNER} & BERT + BiLSTM + CRF + NLP preprocessing & F1\,=\,0.93 & Enhanced BERT with preprocessing excels on noisy social media geographic NER \\
\hline
14 & Manjavacas et al.\ (2026)~\cite{manjavacas2026cvs} & GPT-4o, GPT-OSS-120B, Llama-3.1-8B & GPT-4o $\approx$100\%; Llama $\approx$73\% & Larger models handle complex narrative NER fields better than smaller open-weight models \\
\hline
15 & Stankevi{\v{c}}ius \& Luko{\v{s}}evi{\v{c}}ius (2024)~\cite{stankevivcius2024bert} & BERT-110M, T5, Random Embeddings & Repr.-shaping $\approx$0.70 vs.\ baseline $\approx$0.50 & Simple representation shaping can match complex BERT models on similarity benchmarks \\
\hline
16 & Mansourian \& Oucheikh (2024)~\cite{mansourian2024chatgeoai} & Llama~2 (LangChain + RAG) + NLP extraction & ChatGeoAI $\approx$25\% on complex tasks & LLMs need RAG and entity mapping to execute complex geospatial workflow tasks \\
\hline
17 & Lutsai \& Lampert (2024)~\cite{lutsai2024tweet} & Multilingual BERT + multitask PMOP regression & PMOP: 96\% acc.; 3.68\,km median error & Custom multitask loss with PMOP achieves lowest spatial errors in tweet geolocation \\
\hline
18 & Alsaqer et al.\ (2023)~\cite{alsaqer2023arabic} & LR, SVM, RF, Na\"ive Bayes & Accuracy\,=\,67\% & Classical ML predicts tweet location from text; accuracy limited by short noisy text \\
\hline
19 & Santos et al.\ (2021)~\cite{santos2021address} & BERT-based vs.\ HMM/CRF baselines & BERT F1\,$\approx$\,92--95\%;\newline HMM/CRF\,$\approx$\,80--85\% & BERT significantly outperforms HMM/CRF for parsing unstructured real-world address text \\
\hline
20 & Sun et al.\ (2023)~\cite{sun2023chinesectre} & ChineseCTRE (BERT + BiLSTM-CRF, incr.\ pre-training) & F1\,$\approx$\,91--92\% vs.\ baseline 87--88\% & Incremental domain pre-training with error correction boosts geographic NER recall \\
\hline
21 & Ussenbayeva et al.\ (2026)~\cite{ussenbayeva2026heterogeneous} & XLM-RoBERTa (base), Regex, CRF, Frozen LM & ROUTE\_NUM F1\,=\,0.920;\newline STREET\_NAME F1\,=\,0.282 & Transport NER is heterogeneous; aggregated metrics conceal per-class disparities; composite entities require specialised modelling \\
\hline
\end{longtable}
}

\textbf{Synthesis of literature findings.}A number of common patterns emerged from the literature review that have a significant impact on the design decisions made in this thesis.

\textit{Transformer encoders dominate supervised NER.} In almost all cases where sufficient labeled training data are available, fine-tuning pre-trained transformer based encoders (bert, roberta, XLM-roberta) significantly outperforms both traditional rule-based named entity recognition systems as well as classical statistical models (crf, BiLSTM-CRF). This performance difference is especially dramatic on noisy informal content such as social media or complaints, in which hand-crafted features and gazetteers fail to generalize.

\textit{Multilinguality is non-negotiable for code-switching corpora.} Works~4, 10 and 12 collectively demonstrate that using multilingual models (mbert, XLM-roberta), in addition to being necessary for most real-world applications, provide substantial improvements over their monolingual counterparts when there is mixed language content in the evaluation corpus; regardless of whether the monolingual model was specifically trained for each of the target languages.

\textit{ Competitive Learning Models for Rare Entity Types.} The results in works~1, 2, 8, 9 and 14 demonstrate the instruction-tuned learning models' ability to be most suitable for the extraction of rare entity types. Those entity types typically have fewer examples than other types used during training of fine-tuned encoders. Therefore, there will be overfitting and/or label collapse issues with those fine-tuned encoders. For example, in the CTS dataset, the entity type \texttt{STREET\_INTERSECTION} is only trained by two examples. Thus, LLM's few-shot extraction abilities are the only possible means to extract \texttt{STREET\_INTERSECTION} from the given data.

\textit{Aggregate metrics conceal category-level heterogeneity.} The authors note in their paper~21 (the authors previous research) that the use of a single average F-score can give an overly optimistic view of performance when using a multi-type entity taxonomy. Since the operations costs associated with incorrectly classifying a \texttt{STOP\_NAME}versus not extracting a \texttt{STREET\_INTERSECTION} are different, per-type evaluation is required to accurately evaluate the systems performance and determine if they should be deployed into production.

\textit{Gap analysis.} To date no one has developed a system that addresses all of the following: (a) uses code switching languages (Kazakh & Russian), (b) extracts information from text related to transportation complaints, (c) utilizes a six-type entity classification model including composite space-based entities, and (d) compares four architecturally different methods within the same experimental framework. The authors identify this gap as the basis of the contribution of this thesis.

\section{Setting the task}
\label{sec:task}

\subsection{Formulation of the purpose and objectives of the thesis on automatic geotag recognition}
\label{sec:objectives}

\textbf{Research Aim:} Develop and evaluate a variety of machine learning techniques to automatically recognize geographical locations in a multilingual dataset of text based on complaints about public transportation from two languages (Russian and Kazakh), and assess each method's accuracy and robustness when applied to structured entities with different structures.

\noindent\textbf{Research Object:} The process of identifying names of geographic locations within multilingual texts related to public transportation complaints, which have been collected by City Transportation Systems (CTS) of Astana, Kazakhstan.

\noindent\textbf{Research Subject:} Machine Learning Methods/Algorithms for Automatic Extraction of Six Different Types of Geotag Entities from Noisy, Code-Switching Russian/Kazakh Natural Language Text.

\noindent\textbf{Objectives:}
\begin{enumerate}
  \item Perform an exhaustive comparison of techniques currently used for automatic geotag detection and named entity recognition that can operate on multiple languages and low resource environments.
  \item Gather, process, and label a domain specific multilingual data set of public transportation complaint records with gold standard BIO labels for six different entities.
  \item Develop and apply four machine learning strategies for detecting geotags: Fine-Tuning of pre-trained language models including fine-tuned XLM-RoBERTa, fine-tuned RuBERT, and fine-tuned Multilingual-E5, a few shot LLM Pipeline.
  \item Develop and test four baseline strategies (Regex, Frozen Linear Probe of XLM-R, CRF, Most Frequent Class): to place context around model performance.
  \item Test and Compare All Models Using Per-Class Precision, Recall, F1-Score, Confusion Matrix Analysis, and Boundary Stability Analysis. 
  \item Determine which Model Strategy is Best For Each Entity Type and Overall Which Strategy Would Be Most Suitable For Real-Time Operational Deployment.
\end{enumerate}

\subsection{Requirements for an intelligent geotag-extraction system}
\label{sec:requirements}

The intended deployment context --- real-time triage of CTS complaint records --- imposes the following functional and non-functional requirements on the geotag-extraction system:

\begin{enumerate}
  \item \textbf{Multilingual support.} The system must process Kazakh, Russian, and code-switched sentences without explicit language identification as a prerequisite step.
  \item \textbf{Six-class entity recognition.} The system must classify extracted spans into the six domain-specific entity types: \texttt{ROUTE\_NUM}, \texttt{BUS\_ID}, \texttt{PLATE\_NUM}, \texttt{STOP\_NAME}, \texttt{STREET\_NAME}, \texttt{STREET\_INTERSECTION}.
  \item \textbf{Span-level precision $\geq 0.70$.} Downstream geocoding workflows are sensitive to false positives; high precision is the primary operational constraint to prevent spurious alert generation.
  \item \textbf{Recall coverage of \texttt{STOP\_NAME} and \texttt{ROUTE\_NUM} $\geq 0.80$.} These are the most frequent and operationally critical entities, forming the primary inputs to dispatch routing and hotspot detection.
  \item \textbf{Computational throughput.} For real-time deployment on standard server hardware, the system should be capable of processing at least 100 records per minute, setting a practical limit on model size and inference latency.
  \item \textbf{Graceful degradation on noisy input.} The system must tolerate spelling errors, missing punctuation, non-standard abbreviations, and mixed-script input that are characteristic of citizen complaint text.
  \item \textbf{Structured output.} Extracted entities must be returned as structured data (entity type, extracted text span, character offsets) suitable for downstream GIS linking and database insertion.
\end{enumerate}

\subsection{Formulation of data requirements for model training}
\label{sec:data_requirements}

\textbf{Formal task definition.} We formulate geotag recognition as a token-level sequence labelling task under the BIO (Beginning-Inside-Outside) tagging scheme. Given an input token sequence $\mathbf{T} = (t_1, t_2, \ldots, t_n)$, the objective is to assign each token $t_i$ a label $l_i \in \mathcal{L}$, where:
\begin{equation}
  \mathcal{L} = \{B\text{-}e,\ I\text{-}e \mid e \in \mathcal{E}\} \cup \{O\}, \quad |\mathcal{L}| = 2|\mathcal{E}| + 1 = 13
  \label{eq:labelset}
\end{equation}
$\mathcal{E} = \{\texttt{ROUTE\_NUM}, \texttt{BUS\_ID}, \texttt{PLATE\_NUM}, \texttt{STOP\_NAME}, \texttt{STREET\_NAME}, \texttt{STREET\_INTERSECTION}\}$ denotes the set of six entity types, and $O$ denotes the outside class. The BIO scheme encodes entity spans such that the first token of an entity receives the $B$- prefix (beginning), all subsequent tokens within the same span receive the $I$- prefix (inside), and all non-entity tokens receive $O$ (outside). A predicted entity span $\hat{s}$ is considered a true positive only if both its left and right boundaries exactly match the gold annotation and its entity type label is correct.

\textbf{BIO annotation constraints.} The BIO scheme enforces the following structural constraints on valid label sequences:
\begin{itemize}
  \item An $I\text{-}e$ label must be immediately preceded by either $B\text{-}e$ or $I\text{-}e$ (for the same entity type $e$).
  \item $B\text{-}e$ may follow any label (including $I\text{-}e'$ for a different entity type $e' \neq e$, which terminates $e'$ and begins $e$).
  \item $O$ terminates any open entity span.
\end{itemize}

\textbf{Data requirements.}
\begin{itemize}
  \item A minimum of 2\,000 labelled complaint records to support fine-tuning of transformer encoders without severe overfitting.
  \item Stratified splitting to preserve entity type proportions across training and test sets, ensuring that the test distribution is representative of the full corpus.
  \item Gold-standard annotation produced by at least two independent annotators with native or near-native proficiency in both Kazakh and Russian, with adjudication of disagreements.
  \item Minimum coverage of the rarest entity type: at least 50 training instances of \texttt{STREET\_INTERSECTION} to enable meaningful fine-tuning of minority-class behaviour.
  \item The annotated data must be stored in JSONL format with pre-tokenised BIO label sequences to support efficient DataLoader-based training without on-the-fly tokenisation.
\end{itemize}

\textbf{Class imbalance handling.} Given the severe skew of the entity distribution (Section~\ref{sec:geo_entities}), the following strategies are required to prevent label collapse on minority classes:
\begin{itemize}
  \item \textit{Weighted cross-entropy loss}: minority entity classes receive higher loss weights, computed as $w_c = N_{\text{total}} / (K \cdot N_c)$ where $N_c$ is the count of class $c$.
  \item \textit{Oversampling}: training records containing rare entity types are duplicated $k$ times in the training pool, increasing their effective representation without collecting new data.
  \item \textit{Macro-averaged evaluation}: the primary evaluation metric weights all entity types equally, ensuring that strong performance on frequent classes does not mask failure on rare ones.
\end{itemize}

\subsection{Defining criteria for evaluating the effectiveness of NER models}
\label{sec:eval_criteria}

Model performance is assessed using standard span-level NER metrics computed by the \texttt{seqeval} library, which requires \textit{exact boundary and label match} for a true positive. A predicted span $\hat{s} = (\hat{t}, \hat{l}, \hat{r})$ (type, left boundary, right boundary) is a true positive if and only if there exists a gold span $s^* = (t^*, l^*, r^*)$ such that $\hat{t} = t^*$, $\hat{l} = l^*$, and $\hat{r} = r^*$ simultaneously.

\textbf{Per-class precision, recall, and F1:}
\begin{equation}
  \text{P}_c = \frac{\text{TP}_c}{\text{TP}_c + \text{FP}_c}, \quad
  \text{R}_c = \frac{\text{TP}_c}{\text{TP}_c + \text{FN}_c}, \quad
  F_{1,c} = 2 \cdot \frac{\text{P}_c \cdot \text{R}_c}{\text{P}_c + \text{R}_c}
  \label{eq:per_class_f1}
\end{equation}

\textbf{Macro-averaged F1} weights all entity types equally regardless of frequency:
\begin{equation}
  F_{1,\text{macro}} = \frac{1}{|\mathcal{E}|} \sum_{c \in \mathcal{E}} F_{1,c}
  \label{eq:macro_f1}
\end{equation}
Macro-averaged F1 is the primary evaluation metric because it is sensitive to performance on rare classes such as \texttt{STREET\_INTERSECTION} and \texttt{STREET\_NAME}.

\textbf{Micro-averaged F1} aggregates counts across all classes before computing the ratio:
\begin{equation}
  F_{1,\text{micro}} = 2 \cdot \frac{\sum_c \text{TP}_c}{\sum_c (2\,\text{TP}_c + \text{FP}_c + \text{FN}_c)}
  \label{eq:micro_f1}
\end{equation}
Micro-averaged F1 is influenced primarily by frequent classes and provides an overall measure of throughput-weighted performance.

\textbf{Boundary stability} measures the fraction of correctly labelled predicted entity spans that also have correct boundaries:
\begin{equation}
  \text{BS}_c = \frac{|\{(t, l, r) \in \hat{S}_c : \exists\, (t^*, l^*, r^*) \in S^*_c,\; l = l^* \wedge r = r^*\}|}{|\hat{S}_c \cap S^*_c|_{\text{type-only}}}
\end{equation}
where $|\hat{S}_c \cap S^*_c|_{\text{type-only}}$ counts spans that match in type but not necessarily in boundary. High boundary stability with high precision indicates a model that correctly identifies \textit{what} an entity is but may over- or under-extend its span. Low boundary stability indicates fundamental span segmentation errors that would propagate errors to downstream geocoding.

\textbf{Confusion matrix} analysis at the entity-type level (collapsing BIO prefixes) identifies systematic misclassification patterns. For the CTS entity taxonomy, the most informative confusion axis is between \texttt{BUS\_ID} and \texttt{PLATE\_NUM} (both short alphanumeric identifiers) and between \texttt{STOP\_NAME} and \texttt{STREET\_NAME} (both multi-word spatial toponyms). Confusion between these pairs indicates that the model has learned surface form rather than functional context.

Together, these criteria provide a comprehensive and operationally motivated evaluation framework that goes beyond a single aggregate score and enables category-aware model selection for deployment.