%
% frontmatter/abstract.tex
%
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Geographic information extraction from unstructured textual data has become increasingly critical in domains ranging from smart city management to emergency response systems. This thesis presents a comparative study of machine learning approaches for automatic geotag recognition --- a form of Named Entity Recognition (NER) --- applied to a real-world multilingual dataset of public transport complaint records in Kazakh and Russian languages, provided by City Transportation Systems (CTS) of Astana, Kazakhstan.

Four distinct approaches are developed and evaluated: (1)~fine-tuned XLM-RoBERTa, (2)~fine-tuned ruBERT, (3)~a few-shot Large Language Model pipeline based on Qwen2.5 (7B) deployed locally via Ollama, and (4)~a fine-tuned Multilingual-E5-base (Me5-base) encoder. The dataset comprises 2{,}750 manually annotated complaint records with six domain-specific entity types: \texttt{ROUTE\_NUM}, \texttt{BUS\_ID}, \texttt{PLATE\_NUM}, \texttt{STOP\_NAME}, \texttt{STREET\_NAME}, and \texttt{STREET\_INTERSECTION}, annotated using the BIO tagging scheme.

Experimental results demonstrate that the Qwen2.5 LLM achieves the highest overall precision (0.788) and macro F1-score (0.635), and is the only model capable of detecting the complex \texttt{STREET\_INTERSECTION} entity. Me5-base achieves the best recall (0.653). XLM-RoBERTa provides the strongest single-entity performance on \texttt{ROUTE\_NUM} (F1~=~0.920), while ruBERT (F1~=~0.223) is shown to be inadequate for code-switching multilingual data. These findings provide practical guidance for deploying geotag recognition systems in low-resource, multilingual operational environments.

\bigskip
\keywords{Named Entity Recognition, Geotag Extraction, XLM-RoBERTa, BERT, Large Language Models, Multilingual NLP, Kazakh, Russian, Transport Complaints, BIO Tagging, Sequence Labeling}

\clearpage
