%
% frontmatter/intro.tex
%

\chapter*{Introduction}
\phantomsection
\addcontentsline{toc}{chapter}{Introduction}

The increasing amount of digitalized data related to transportation networks is producing an increasing number of complaints from citizens. Because of this, public transport entities are receiving a large amount of complaint messages on a daily basis through mobile apps, web sites, and phone calls. In each message or complaint, a passenger provides their own account of an operational failure; i.e., they describe how the operation failed (e.g. a bus missed a scheduled stop, the driver deviated from the route, etc.) using free form text. Within this unformatted text, there are many geographic locations mentioned -- e.g. route numbers, names of bus stops, names of streets, identifiers for vehicles and intersections, etc. -- that could be used by various transportation network operators to improve real-time monitoring of service performance, optimize routes, and automate response to incidents. Thus, as both the volumes of input grow and the costs associated with manually reviewing every complaint continue to increase, there is a compelling business case for developing automated tools to identify relevant geographic information from complaint messages.

Although there are many studies that focus on named-entity recognition (NER), most NER models were trained using high-resource languages and generic domains like news articles or biomedicine. The reason why the majority of existing models do not provide sufficient detail for transport-domain operation texts from low resource bilingual settings is that they have not been precisely customized for them. The CTS complaints in Astana, Kazakhstan represent an example of this. There are approximately 3000 monthly complaint records processed by City Transportation System (CTS). However, all complaints were written in a language mix of Kazakh and Russian and had multiple instances of code-switching; informal register; numerous abbreviations relevant to transportation; and varying levels of punctuation. To date, none of the previous works targeted all these specific combinations: Kazakh-Russian code-switching; transport-domain complaint text; a six-class entity classification system, which includes several types of structurally complex composite entities; and a comparison of different machine-learning architectures applied to the same data under equivalent experimental conditions.

To meet these requirements, this study intends to develop and compare several machine learning techniques for identifying geotags automatically using multi-language textual information. As such, it is intended that the results of this study will be used as part of an operational solution for CTS.

This is achieved through the development of the study, which includes (i) examining the existing literature on named entity recognition and geotagging in relation to four paradigms -- rule-based, statistical, neural and large language model -- (ii) creating a dataset of 2\,750 annotated public transport complaints in Kazakh and Russian languages (including pre-processing); (iii) implementing and evaluating four machine learning architectures -- including three based on a pre-trained XLM-RoBERTA, two based on pre-trained DeepPavlov ruBERT, one based on Multilingual-E5-base and one based on few-shot Qwen2.5 LLM inference -- and their corresponding baselines relative to precision, recall and F1 scores for each class and globally; and (iv) integrating the best performing architecture into an optimal inference pipeline for use in real time CTS complaint handling. 
The object of investigation is the named entity recognition within the context of public transportation complaint texts written in both Kazakh and Russian languages; the subject of the investigation is the family of machine learning models and algorithms capable of extracting six different geotag entities from noisy, code-switching public transportation complaint texts.

Chapter~1. Explanation of theoretical bases for geographic named entities in transport domain text through analysis of linguistic and operational features. This chapter summarises current approaches to automated entity recognition from rule-based, statistical methods, transformer encoders and large language models, as well as particular difficulties of Kazakh-Russian code-switching and morphological complexity. Furthermore it provides a brief summary of 21 key related works. On this basis, the chapter determines the objective of the study, defines six types of entitiy, sets out requirements for training data and formulates quantative criteria for evaluating effectiveness of model.

Chapter~2. The chapter covers model development and experimental configuration. The section outlines the methodology for collecting, annotating, and pre-processing the data; it explains how the code-switched data will be pre-processed as well as sub-word aligned for use in the four major models. It provides the details regarding the training parameters used with each of the four major models. It describes the four baselines. Finally, it discusses the metrics that were used for evaluating all of the models.

Chapter~3. This chapter is the first of two chapters that provide an overview of the results from the comparative experiments conducted on this study. The chapter includes tables which summarize the performance of each of the four models being evaluated. In addition to providing those summary tables, the chapter also contains a description of the errors encountered during testing (e.g., rare entities; failed code-switches; noisy input) by each of the models. Finally, since multiple models are being compared within this study, there is also an overall cross-model comparison in order to provide a basis for recommending a single model for future use in the CTS production environment.

Beginning with Chapter~4, the main findings are summarized; the contribution of this study is outlined as well as potential avenues for further investigation, namely; (1) augmenting datasets with targeted data regarding rare entities, and (2), developing hybrid architecture systems that incorporate both encoders and LLMs.

This work has an applied or practical value due to its ability to directly address the needs of real world public transportation managers. The proposed system enables transport managers to rapidly and accurately perform geotagging on citizen complaint reports. This enables them to respond to incidents quickly, identify geographic "hot spots" automatically and to diminish their reliance upon manually processing these complaints by trained staff. Additionally, the annotated dataset and comparative results will serve as a starting point for future studies in which multilingual NLP meets smart city development.